Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-04 LaTeX/Turnover.tex
generic loci order in Matt version analysis

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Therefore, the invasion of rare dominant neo-Y or neo-W alleles depends on only two equations and is therefore determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $2\lambda_{mi}-1$ is the growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination in heterozygotes, and $\rho_{mi}-1$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination in heterozygotes (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations as there are weird 2s in table 2}

\begin{threeparttable}[ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{Parameters determining invasion of neo-Y or neo-W alleles into ancestral XY system}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{X}^{\female} w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male}+2(1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male}\alpha^\male(1-R)\}/ \{ 2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male\} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{X}^{\female}) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male}+2p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1-\alpha^\male)(1-R)\}/ \{ 2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} \alpha^{\male}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha^{\male})/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+2(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}\alpha^\female(1-R)\}/ \{ 2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+2\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha^\female)(1-R)\}/ \{2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} \alpha^{\female}/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha^{\female})/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item $\bar{p}^{\male}=p_{Y}^{\male}q+p_{X}^{\male}(1-q)$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
      \item See Table \ref{tab:meanfitnesses} for expressions of mean fitnesses.
%      \item $\bar{w}_H^\female = p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ is the mean fitness of female gametes during gametic competition
%      \item $\bar{w}_H^\male = \bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ is the mean fitness of male gametes during gametic competition
%       \item $\bar{w}^\female = \{ p_X^\female w_A^\female (1-q) p_X^\male w_A^\male w_{AA}^\female + (1 - p_X^\female) w_a^\female (1-q) p_X^\male w_A^\male w_{Aa}^\female + p_X^\female w_A^\female (1-q) (1 - p_X^\male) w_a^\male w_{Aa}^\female + (1-p_X^\female) w_a^\female (1-q) (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}$ is the mean fitness of females during diploid competition
%       \item $\bar{w}^\male = \{ p_X^\female w_A^\female q p_Y^\male w_A^\male w_{AA}^\male + (1 - p_X^\female) w_a^\female q p_Y^\male w_A^\male w_{Aa}^\male + p_X^\female w_A^\female q (1 - p_Y^\male) w_a^\male w_{Aa}^\male + (1-p_X^\female) w_a^\female q (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\}$ is the mean fitness of males during diploid competition
    \end{tablenotes}
  \label{tab:haplotype_growth}
\end{threeparttable}
\\

@@

We are particularly concerned with whether or not a rare neo-sex-determining allele increases in frequency, which occurs when the largest eigenvalue, $\lambda$, is greater than one. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (i.e., if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2 > 0$ then $\lambda > 1$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ 
\frac{p^\female_X w^\female_A w^\male_a (1-\alpha^\male)}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{mA}-1)} + 
%p^\female_X \frac{w^\female_A}{\bar{w}^\female_H} \frac{w^\male_a(1-\alpha^\male)}{\bar{w}^\male_H} \frac{1}{\lambda_{mA}-1} + 
\frac{(1 - p^\female_X) w^\female_a w^\male_A \alpha^\male}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{ma}-1)} 
\right] 
\frac{w^\male_{Aa}}{\bar{w}^\male} < 1,
\end{equation}

@@

\noindent
for the neo-$W$. Equations \eqref{eq:neoYR} and \eqref{eq:neoWR} show that the new sex-determining allele, $m$, is expected to invade for any recombination rate, $R$, when the net flow of recombinants is from the less fit (smaller $\lambda_{mi}$) to the more fit \textbf{A} background (making the terms inside the square brackets in Equations \ref{eq:neoYR} and \ref{eq:neoWR} negative). \textcolor{red}{Q: is it definitely possible to have negative square brackets for a equilibria maintained by selection?}
%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^\female_X) w_{m,A} g_a^* < p^\female_X w_{m,a} g_A^*$.
When the net flow of recombinants is from the more fit to the less fit haplotype, the new sex-determining allele can still invade when the rate of recombination between it and the selected locus, $R$, is small enough. \textcolor{red}{Q:Is it the case that sometimes the square brackets are positive and invasion occurs for $R=1/2$? In which case it might be better to have slightly different phrasing here.}

>>>>

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Therefore, the invasion of rare dominant neo-Y or neo-W alleles depends on only two equations and is therefore determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $2\lambda_{mi}-1$ is the growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}-1$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females? the two in the numerator is from heterozygotes (like Hardy-Weinberg)?}

\begin{threeparttable}[ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{X}^{\female} w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + 2(1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} \alpha^\male (1 - r_{AM})\}/ \{ 2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male\} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{X}^{\female}) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + 2p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha^\male) (1 - r_{AM})\}/ \{ 2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=r_{AM} (1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} \alpha^{\male}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=r_{AM} p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha^{\male})/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+2(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}\alpha^\female(1-R)\}/ \{ 2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+2\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha^\female)(1-R)\}/ \{2 \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} \alpha^{\female}/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha^{\female})/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item $\bar{p}^{\male}=p_{Y}^{\male}q+p_{X}^{\male}(1-q)$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
      \item $r_{AM}$ is the probability of recombination between loci \textbf{A} and \textbf{M}.
      \item See Table \ref{tab:meanfitnesses} for expressions of mean fitnesses.
%      \item $\bar{w}_H^\female = p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ is the mean fitness of female gametes during gametic competition
%      \item $\bar{w}_H^\male = \bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ is the mean fitness of male gametes during gametic competition
%       \item $\bar{w}^\female = \{ p_X^\female w_A^\female (1-q) p_X^\male w_A^\male w_{AA}^\female + (1 - p_X^\female) w_a^\female (1-q) p_X^\male w_A^\male w_{Aa}^\female + p_X^\female w_A^\female (1-q) (1 - p_X^\male) w_a^\male w_{Aa}^\female + (1-p_X^\female) w_a^\female (1-q) (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}$ is the mean fitness of females during diploid competition
%       \item $\bar{w}^\male = \{ p_X^\female w_A^\female q p_Y^\male w_A^\male w_{AA}^\male + (1 - p_X^\female) w_a^\female q p_Y^\male w_A^\male w_{Aa}^\male + p_X^\female w_A^\female q (1 - p_Y^\male) w_a^\male w_{Aa}^\male + (1-p_X^\female) w_a^\female q (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\}$ is the mean fitness of males during diploid competition
    \end{tablenotes}
  \label{tab:haplotype_growth}
\end{threeparttable}
\\

@@

We are particularly concerned with whether or not a rare neo-sex-determining allele increases in frequency, which occurs when the largest eigenvalue, $\lambda$, is greater than one. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (i.e., if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2 > 0$ then $\lambda > 1$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
r_{AM} \left[ 
\frac{p^\female_X w^\female_A w^\male_a (1-\alpha^\male)}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{mA}-1)} + 
%p^\female_X \frac{w^\female_A}{\bar{w}^\female_H} \frac{w^\male_a(1-\alpha^\male)}{\bar{w}^\male_H} \frac{1}{\lambda_{mA}-1} + 
\frac{(1 - p^\female_X) w^\female_a w^\male_A \alpha^\male}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{ma}-1)} 
\right] 
\frac{w^\male_{Aa}}{\bar{w}^\male} < 1,
\end{equation}

@@

\noindent
for the neo-$W$. Equations \eqref{eq:neoYR} and \eqref{eq:neoWR} show that the new sex-determining allele, $m$, is expected to invade for any probability of recombination between loci \textbf{A} and \textbf{M}, $r_{AM}$, when the net flow of recombinants is from the less fit (smaller $\lambda_{mi}$) to the more fit \textbf{A} background (making the terms inside the square brackets in Equations \ref{eq:neoYR} and \ref{eq:neoWR} negative). \textcolor{red}{Q: is it definitely possible to have negative square brackets for a equilibria maintained by selection?}
%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^\female_X) w_{m,A} g_a^* < p^\female_X w_{m,a} g_A^*$.
When the net flow of recombinants is from the more fit to the less fit haplotype, the new sex-determining allele can still invade when the rate of recombination between it and the selected locus is small enough. \textcolor{red}{Q:Is it the case that sometimes the square brackets are positive and invasion occurs for $r_{AM}=1/2$? In which case it might be better to have slightly different phrasing here.}
